---
title: "outlierApproach"
author: "tatum katz"
date: "2023-09-22"
output: html_document
---

# run this script after running cleaning.Rmd to perform the outlier approach to identify serotypes of greater concern

# libraries

```{r libraries}

library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(lubridate)
library(MASS)
library(pROC)
library(car)
library(cluster)
library(factoextra)
library(dendextend)
library(purrr)
library(ggrepel)
library(outliers)
library(kableExtra)

```

# data

```{r data}

nors <- read.csv("nors.csv")

```

## make ratio dataset

```{r}
ratioDat <- nors %>%
  filter(meatSource != "multiple") %>% # remove outbreaks caused by multiple CAFC sources
  drop_na(HospitalInfo, DeathsInfo) %>%
  group_by(serotype, meatSource, IFSACLevel4_1) %>% # level 4 gives the commodities we want
  dplyr::summarize(nOut = n(), # get total number of each event
            nIll=sum(EstimatedPrimary),
            nHosp=sum(HospitalNum),nHospInfo=sum(HospitalInfo),
            nDeath=sum(DeathsNum), nDeathInfo=sum(DeathsInfo)) %>% # spread the data out first
  pivot_wider(id_cols=c(serotype, IFSACLevel4_1), # total number of each event, per source (meat, nonmeat, multiple, NA)
              names_from=c(meatSource),
              values_from=c(nOut, nIll, nHosp, nHospInfo, nDeath, nDeathInfo)) %>%
  pivot_wider(id_cols=c(serotype), # total number of each event, per IFSAC level 4
              names_from=c(IFSACLevel4_1),
              values_from=grep("^n", names(.), value=T)) %>%
  replace(is.na(.),0) %>% # fill all NAs with zero because NA means 0 in this dataset and we want to calculate lots of stuff
  dplyr::mutate( # now calculate the values we want
          # total trials
         nOutTotal = sum(c_across(starts_with("nOut"))), # total number of outbreaks caused by each sero across all sources
         nIllTotal = sum(c_across(starts_with("nIll_"))), # total number of illnesses caused by each sero across all sources
         meanOutSize = nIllTotal/nOutTotal, # mean outbreak size
         nHospTotal = sum(c_across(starts_with("nHosp_"))), # total number of hospitalizations caused by each sero across all sources
          # total successes - meatSource (meat)
         nOutMeat = sum(c_across(starts_with("nOut_meat_"))), # total number of outbreaks caused by each sero from meat sources
         nIllMeat = sum(c_across(starts_with("nIll_meat_"))), # total number of illnesses caused by each sero from meat sources
         nHospMeat = sum(c_across(starts_with("nHosp_meat_"))), # total number of hospitalizations caused by each sero from meat sources
          # total successes - beef, chicken, pork, turkey
         nOutBeef = nOut_meat_Beef, # number of outbreaks caused by each sero from beef
         nOutChicken = nOut_meat_Chicken, # number of outbreaks caused by each sero from chicken
         nOutPork = nOut_meat_Pork, # number of outbreaks caused by each sero from pork
         nOutTurkey = nOut_meat_Turkey, # number of outbreaks caused by each sero from turkey
         nIllBeef = nIll_meat_Beef, # number of ill caused by each sero from beef
         nIllChicken = nIll_meat_Chicken, # number of ill caused by each sero from chicken
         nIllPork = nIll_meat_Pork, # number of ill caused by each sero from pork
         nIllTurkey = nIll_meat_Turkey, # number of ill caused by each sero from turkey
         meanOutSizeBeef = nIll_meat_Beef/nOut_meat_Beef,
         meanOutSizeChicken = nIll_meat_Chicken/nOut_meat_Chicken,
         meanOutSizePork = nIll_meat_Pork/nOut_meat_Pork,
         meanOutSizeTurkey = nIll_meat_Turkey/nOut_meat_Turkey,
         nHospBeef = nHosp_meat_Beef, # number of hosp caused by each sero from beef
         nHospChicken = nHosp_meat_Chicken, # number of hosp caused by each sero from chicken
         nHospPork = nHosp_meat_Pork, # number of hosp caused by each sero from pork
         nHospTurkey = nHosp_meat_Turkey, # number of hosp caused by each sero from turkey
          # ratio of hospitalizations:illnesses (all meat)
         ratioHospIllMeat = nHospMeat / nIllMeat,
          # ratio of hospitalizations:outbreaks (beef)
         ratioHospIllBeef = nHospBeef / nIllBeef,
          # ratio of hospitalizations:outbreaks (chicken)
         ratioHospIllChicken = nHospChicken / nIllChicken,
          # ratio of hospitalizations:outbreaks (pork)
         ratioHospIllPork = nHospPork / nIllPork,
          # ratio of hospitalizations:outbreaks (turkey)
         ratioHospIllTurkey = nHospTurkey / nIllTurkey,
         ) %>%
  dplyr::select(!grep("_", names(.), value=T)) %>% # remove necessary columns, basically any column with an _
  dplyr::mutate(across(.cols = everything(), ~ ifelse(is.infinite(.x) | is.na(.x) | is.nan(.x), 
                                               0, .x))) # turn all NAs and Infs to zeros (because there were 0 of these events)

```

# IQR outlier detection

IQR: "all values lower than the first and greater than the third quartile is considered, and the difference between them and the nearest quartile divided by the IQR are calculated. values between these quartiles are given a score of 0" - no pvalues will be calculated

formula: value - Q3(or Q1) / IQR  

so if the outcome is greater than one, the difference between the value and the quartile value is GREATER than the IQR

```{r}

## meat overall ##
# number of outbreaks
meatIQR <- data.frame(serotype=ratioDat$serotype, 
                      iqrOutbreaks = scores(ratioDat$meanOutSize, type = "iqr"),
                      limOutbreaks = scores(ratioDat$meanOutSize, type = "iqr", lim=1),
                      iqrHospToIll = scores(ratioDat$ratioHospIllMeat, type="iqr"),
                      limHospToIll = scores(ratioDat$ratioHospIllMeat, type="iqr", lim=1))

# which serotypes are outliers by iqr?
meatIQR[which(meatIQR$limOutbreaks==T),] # for outbreak size - blockley carrau heidelberg montevideo newport oranienburg poona reading
meatIQR[which(meatIQR$limHospToIll==T),] # for ratio - dublin, eastbourne, javiana, paratyphiB, reading, unknown
meatIQR[which(meatIQR$limOutbreaks==T & meatIQR$limHospToIll==T),] # outliers in both - reading only

## beef ##
# create results dataframe
beefIQR <- data.frame(serotype=ratioDat$serotype, 
                      iqrOutbreaks = scores(ratioDat$meanOutSizeBeef, type = "iqr"), 
                      limOutbreaks = scores(ratioDat$meanOutSizeBeef, type = "iqr", lim=1),
                      iqrHospToIll = scores(ratioDat$ratioHospIllBeef, type="iqr"),
                      limHospToIll = scores(ratioDat$ratioHospIllBeef, type="iqr", lim=1))

# which serotypes are outliers by iqr?
beefIQR[which(beefIQR$limOutbreaks==T),] # for outbreak size - Braenderup, Dublin, Enteritidis, Heidelberg, I,4,[5],12:i:-, Infantis, Javiana, Montevideo, Muenchen, Newport, Oranienburg, Potsdam, Typhimurium, Uganda
beefIQR[which(beefIQR$limHospToIll==T),] # for ratio - branderup, dublin, enteritidis, heidelberg, i4512i-, infantis, monte, muenchen, newport, oranienburg, potsdam, typhimurium, uganda
beefIQR[which(beefIQR$limOutbreaks==T & beefIQR$limHospToIll==T),] # outliers in both -  branderup, dublin, enteritidis, heidelberg, i4512i-, infantis, monte, muenchen, newport, oranienburg, potsdam, typhimurium, uganda


## chicken ##
# create results dataframe
chickenIQR <- data.frame(serotype=ratioDat$serotype, 
                      iqrOutbreaks = scores(ratioDat$meanOutSizeChicken, type = "iqr"), 
                      limOutbreaks = scores(ratioDat$meanOutSizeChicken, type = "iqr", lim=1), 
                      iqrHospToIll = scores(ratioDat$ratioHospIllChicken, type="iqr"),
                      limHospToIll = scores(ratioDat$ratioHospIllChicken, type="iqr", lim=1))

# which serotypes are outliers by iqr?
chickenIQR[which(chickenIQR$limOutbreaks==T),] # for noutbreak size - blockley, braenderup, carmel, enteritidis, heidelberg, i4512i-, infantis, javiana, montevideo, saintpaul, schwarzengrund, thompson, typhimurium
chickenIQR[which(chickenIQR$limHospToIll==T),] # for ratio - blockley, branderup, carmel, heidelberg, i4512i-, infantis, javiana, montevideo, newport, norwich, paratyphiB, saintpaul, thompson, typhimurium
chickenIQR[which(chickenIQR$limOutbreaks==T & chickenIQR$limHospToIll==T),] # outliers in both - blockley, braenderup, carmel, heidelberg, i4512i-, infantis, javiana, montevideo, saintpaul, thompson, typhimurum


## pork ##
# create results dataframe
porkIQR <- data.frame(serotype=ratioDat$serotype, 
                      iqrOutbreaks = scores(ratioDat$meanOutSizePork, type = "iqr"), 
                      limOutbreaks = scores(ratioDat$meanOutSizePork, type = "iqr", lim=1), 
                      iqrHospToIll = scores(ratioDat$ratioHospIllPork, type="iqr"),
                      limHospToIll = scores(ratioDat$ratioHospIllPork, type="iqr", lim=1)) 

# which serotypes are outliers by iqr?
porkIQR[which(porkIQR$limOutbreaks==T),] # for outbreak size - berta, give, i4512i-, infantis, javiana, muenchen, schwarzengrund, typhimurium
porkIQR[which(porkIQR$limHospToIll==T),] # for ratio - berta, derby, eastbourne, i4512i-, infantis, newport, schwarzengrund, thompson, uganda, unknown
porkIQR[which(porkIQR$limOutbreaks==T & porkIQR$limHospToIll==T),] # outliers in both - berta, i4512i-, infantis, schwarzengrund

## turkey ##
# create results dataframe
turkeyIQR <- data.frame(serotype=ratioDat$serotype, 
                        iqrOutbreaks = scores(ratioDat$meanOutSizeTurkey, type = "iqr"), 
                        limOutbreaks = scores(ratioDat$meanOutSizeTurkey, type = "iqr", lim=1), 
                        iqrHospToIll = scores(ratioDat$ratioHospIllTurkey, type="iqr"),
                        limHospToIll = scores(ratioDat$ratioHospIllTurkey, type="iqr", lim=1))

# which serotypes are outliers by iqr?
turkeyIQR[which(turkeyIQR$limOutbreaks==T),] # for n outbreaks - anatum, berta, braenderup, enteritidis, hadar, heidelberg, i4512i-, javiana, muenchen, newport, reading, saintpaul, schwarzengrund, typhimurium
turkeyIQR[which(turkeyIQR$limHospToIll==T),] # for ratio - anatum, berta, enteritidis, hadar, heidelberg, javiana, muenchen, newport, reading, saintpaul, schwarzengrund, typhimurium
turkeyIQR[which(turkeyIQR$limOutbreaks==T & turkeyIQR$limHospToIll==T),] # outliers in both - anatum, berta, enteritidis, hadar, heidelberg, javiana, muenchen, newport, reading, saintpaul, schwarzengrund, typhimurium

```

# outputs for supplementary materials

first calculate outlier cutoffs

```{r}

# remember score cutoff is anything greater than 1
# find third quartile
# find IQR
# x < IQR + Q3

meatOutCut <- unname(IQR(ratioDat$meanOutSize)+quantile(ratioDat$meanOutSize, 3/4)) # >60.60
meatRatCut <- unname(IQR(ratioDat$ratioHospIllMeat)+quantile(ratioDat$ratioHospIllMeat, 3/4)) # >.30

beefOutCut <- unname(IQR(ratioDat$meanOutSizeBeef)+quantile(ratioDat$meanOutSizeBeef, 3/4)) # >0
beefRatCut <- unname(IQR(ratioDat$ratioHospIllBeef)+quantile(ratioDat$ratioHospIllBeef, 3/4)) # >0

chickenOutCut <- unname(IQR(ratioDat$meanOutSizeChicken)+quantile(ratioDat$meanOutSizeChicken, 3/4)) # >14.5
chickenRatCut <- unname(IQR(ratioDat$ratioHospIllChicken)+quantile(ratioDat$ratioHospIllChicken, 3/4)) # >0

porkOutCut <- unname(IQR(ratioDat$meanOutSizePork)+quantile(ratioDat$meanOutSizePork, 3/4)) # >24.5
porkRatCut <- unname(IQR(ratioDat$ratioHospIllPork)+quantile(ratioDat$ratioHospIllPork, 3/4)) # >0.15

turkeyOutCut <- unname(IQR(ratioDat$meanOutSizeTurkey)+quantile(ratioDat$meanOutSizeTurkey, 3/4)) # >0
turkeyRatCut <- unname(IQR(ratioDat$ratioHospIllTurkey)+quantile(ratioDat$ratioHospIllTurkey, 3/4)) # >0

meatRules <- c(meatRatCut, meatOutCut, paste0(iqrMeat, collapse=", "))
beefRules <- c(beefRatCut, beefOutCut, paste0(iqrBeef, collapse=", "))
chickenRules <- c(chickenRatCut,chickenOutCut, paste0(iqrChick, collapse=", "))
porkRules <- c(porkRatCut, porkOutCut, paste0(iqrPork, collapse=", "))
turkeyRules <- c(turkeyRatCut, turkeyOutCut, paste0(iqrTurk, collapse=", "))

rulesTab <- as.data.frame(cbind(meatRules, beefRules, chickenRules, porkRules, turkeyRules), row.names = c("Hospitalizations to Illness Ratio", "Average Number of Outbreaks", "Serotypes of Concern"))
colnames(rulesTab) <- c("Meat Overall", "Beef", "Chicken", "Pork", "Turkey")

rulesTab %>% kbl # generate the results table

```

make figures

```{r}

m <- meatIQR %>% 
  mutate(noOutbreaks = ratioDat$nOutMeat,
         ratioHospIll = ratioDat$ratioHospIllMeat,
         colorcode = ifelse(limOutbreaks==T, ifelse(limHospToIll==T, "both", "outOnly"), ifelse(limHospToIll==T, "ratioOnly", "neither")),
         label = ifelse(limOutbreaks==T | limHospToIll==T, serotype, "")) %>%
ggplot(aes(x=noOutbreaks, y=ratioHospIll, color=colorcode))+
  geom_point()+
  scale_color_manual(values=c("purple", "red", "blue", "black"),
                     breaks=c("both", "outOnly", "ratioOnly", "neither"),
                     labels=c("Outlied in Both", "Outlied in Outbreaks", "Outlied in Ratio", "Not an Outlier"))+
  geom_hline(yintercept=meatRatCut, color="blue")+
  geom_vline(xintercept=meatOutCut, color="red")+
  labs(x="Outbreaks",
       y="Hospitalizations : Illnesses",
       title="Meat Overall",
       color=" ")+
  geom_label_repel(aes(label=label), size=1.5, min.segment.length=0.1, max.overlaps=100, show.legend=F)+
  theme_bw()+
  theme(legend.position="none")

b <- beefIQR %>% 
  mutate(noOutbreaks = ratioDat$nOutBeef,
         ratioHospIll = ratioDat$ratioHospIllBeef,
         colorcode = ifelse(limOutbreaks==T, ifelse(limHospToIll==T, "both", "outOnly"), ifelse(limHospToIll==T, "ratioOnly", "neither")),
                  label = ifelse(limOutbreaks==T | limHospToIll==T, serotype, "")) %>%
ggplot(aes(x=noOutbreaks, y=ratioHospIll, color=colorcode))+
  geom_point()+
  scale_color_manual(values=c("purple", "red", "blue", "black"),
                     breaks=c("both", "outOnly", "ratioOnly", "neither"),
                     labels=c("Outlied in Both", "Outlied in Outbreaks", "Outlied in Ratio", "Not an Outlier"))+
  geom_hline(yintercept=beefRatCut, color="blue")+
  geom_vline(xintercept=beefOutCut, color="red")+
  labs(x="Outbreaks",
       y="Hospitalizations : Illnesses",
       title="Beef",
       color=" ")+
  geom_label_repel(aes(label=label), size=1.5, max.overlaps=100, min.segment.length=0.1, show.legend=F)+
  theme_bw()+
  theme(legend.position="none")

c <- chickenIQR %>% 
  mutate(noOutbreaks = ratioDat$nOutChicken,
         ratioHospIll = ratioDat$ratioHospIllChicken,
         colorcode = ifelse(limOutbreaks==T, ifelse(limHospToIll==T, "both", "outOnly"), ifelse(limHospToIll==T, "ratioOnly", "neither")),
         label = ifelse(limOutbreaks==T | limHospToIll==T, serotype, "")) %>%
ggplot(aes(x=noOutbreaks, y=ratioHospIll, color=colorcode))+
  geom_point()+
  scale_color_manual(values=c("purple", "red", "blue", "black"),
                     breaks=c("both", "outOnly", "ratioOnly", "neither"),
                     labels=c("Outlied in Both", "Outlied in Outbreaks", "Outlied in Ratio", "Not an Outlier"))+
  geom_hline(yintercept=chickenRatCut, color="blue")+
  geom_vline(xintercept=chickenOutCut, color="red")+
  labs(x="Outbreaks",
       y="Hospitalizations : Illnesses",
       title="Chicken",
       color=" ")+
  geom_label_repel(aes(label=label), size=1.5, min.segment.length=0.1, max.overlaps=100, show.legend=F)+
  theme_bw()+
  theme(legend.position="none")

p <- porkIQR %>% 
  mutate(noOutbreaks = ratioDat$nOutPork,
         ratioHospIll = ratioDat$ratioHospIllPork,
         colorcode = ifelse(limOutbreaks==T, ifelse(limHospToIll==T, "both", "outOnly"), ifelse(limHospToIll==T, "ratioOnly", "neither")),
         label = ifelse(limOutbreaks==T | limHospToIll==T, serotype, "")) %>%
ggplot(aes(x=noOutbreaks, y=ratioHospIll, color=colorcode))+
  geom_point()+
  scale_color_manual(values=c("purple", "red", "blue", "black"),
                     breaks=c("both", "outOnly", "ratioOnly", "neither"),
                     labels=c("Outlied in Both", "Outlied in Outbreaks", "Outlied in Ratio", "Not an Outlier"))+
  geom_hline(yintercept=porkRatCut, color="blue")+
  geom_vline(xintercept=porkOutCut, color="red")+
  labs(x="Outbreaks",
       y="Hospitalizations : Illnesses",
       title="Pork",
       color=" ")+
  geom_label_repel(aes(label=label), size=1.5, max.overlaps=100,  min.segment.length=0.1, show.legend=F)+
  theme_bw()

t <- turkeyIQR %>% 
  mutate(noOutbreaks = ratioDat$nOutTurkey,
         ratioHospIll = ratioDat$ratioHospIllTurkey,
         colorcode = ifelse(limOutbreaks==T, ifelse(limHospToIll==T, "both", "outOnly"), ifelse(limHospToIll==T, "ratioOnly", "neither")),
         label = ifelse(limOutbreaks==T | limHospToIll==T, serotype, "")) %>%
ggplot(aes(x=noOutbreaks, y=ratioHospIll, color=colorcode))+
  geom_point()+
  scale_color_manual(values=c("purple", "red", "blue", "black"),
                     breaks=c("both", "outOnly", "ratioOnly", "neither"),
                     labels=c("Outlied in Both", "Outlied in Outbreaks", "Outlied in Ratio", "Not an Outlier"))+
  geom_hline(yintercept=turkeyRatCut, color="blue")+
  geom_vline(xintercept=turkeyOutCut, color="red")+
  labs(x="Outbreaks",
       y="Hospitalizations : Illnesses",
       title="Turkey",
       color=" ")+
  geom_label_repel(aes(label=label), size=1.5, max.overlaps=100,  min.segment.length=0.1, show.legend=F)+
  theme_bw()+
  theme(legend.position="none")

ggarrange(m, b, c, 
          p+theme(legend.position="none"), t, get_legend(p),
          nrow=2, ncol=3)

```
